---
title: "R Notebook Tutorial"
author: "Eric R. Scott"
date: '2018-01-16'
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
subtitle: Tufts BUGS workshop
---
# What is an R Notebook?
 
 - A file format (.rmd) that can be opened by R Studio
 - Allow you to mix prose, code, and code output
 - Uses a simple system called Markdown for formatting prose
 - Markdown headings allow easy navigation in R Studio
 - Notebooks can be "knit" to produce beautiful reports as html, pdf, or Word documents
 - Allows configuration of what to show in these outputs (e.g. code, output, warnings)

# Why Notebooks?

 - It's a good practice to annotate your code anyway, and .rmd lets you do this in a richer, easier to read way
 - As an analysis notebook.  Organizing your code into "chunks" interspersed with prose helps you think through problems better and keep track of your thought process.
 - Portability/collaboration. No need to use `setwd()`
 - Beautiful, easy to navigate reports. You can hide or show code depending on who the report is for (e.g. does this person know R?)
 - Creating examples, tutorials, assignments, etc. using R
 - Plain text (despite richly formatted output), so very compatible with Git and other version control.
 
# Workshop Outline

## Make a Notebook!

- Make a new notebook, run the chunks, preview the output
- avoid spaces in filename
    
## Reports

- "knit" button calls functions from `knitr` package
- pdf generation requires LaTeX installation
- difference between "Preview" and "knit..." is preview does not run all chunks explicitly.  Knit fails if code doesn't run.

## Basics of Markdown

### Basic Formatting

- **bold**
- *itallics*
- subscript^x^
- superscript~y~
- formatting as `code` with backticks
- lists (must have space between list header and first list item)

### Headings and subheadings

- using different nubmers of "#"
- integration with RStudio interface

### Equations
- Inline with "\$". E.g. n = 5 vs. $n=5$.
- Display equation with "\$\$"
$$
n=5
$$

- More complicated equations can be made using LaTeX:

$$
s = \sqrt{\frac{\sum{}y_i^2 - \frac{(\sum{}y_i)^2}{n}}{n-1}}
$$
- A cheat-sheet for making equations like this can be found [here](https://en.wikibooks.org/wiki/LaTeX/Mathematics)

### When Markdown doesn't work

- R Markdown is "portable".  It works for html, pdf, Word.
- However, if your output is html, you can use html.
- If your output is pdf, you can use LaTeX
- If your output is Word, just edit it in Word
- R Markdown is limited, but you probably won't need super advanced formatting
- Getting help:  
    + *Help > Markdown Quick Reference*
    + *Help > Cheatsheets > R Markdown Cheatsheet*
    + *Help > Cheatsheets > R Markdown Reference Guide*
    + Google

## Code chunks
### Inserting and naming code chunks

- Insert with menu, keyboard shortcut, or just typing
- naming code chunks and integration with R Studio (**Duplicate names not allowed**)
- Inline code with \` r #code \`

### Running code in code chunks

- "play" button
- one line at a time with *cmd + enter*
- run all previous with down button.
- "Run" button.  Use *Restart R and Runn All Chunks* to test self-containedness.

- Self-containedness of R Notebooks
    + don't use `setwd()`
    + use absolute paths.  Also package `here()` might be helpful to make things more self-contained
    + shareable (in html output there is even a download code button)
    + 

- If your code chunk produces multiple outputs, they will show up in separate "tabs."  
```{r eval=FALSE, include=TRUE}
plot(cars)
plot(trees)
cor(trees)
```
- in pdf or word, output appears after code that generated it.

### Code chunk options

 - gear icon to access.
 - specify what shows up in output. (useful for sharing results with non-R-literate)
 - don't show errors or warnings (useful for package loading chunks)
 - don't run code (useful for keeping broken code visible.  Remember, won't knit if code throws errors.)
 - figure size specification (**Note**: Base R plots have way larger margins than ggplot.  Be careful when mixing.)
 
```{r eval=FALSE, fig.height=2, fig.width=2, include=TRUE}
#base R plot resized to 2" x 2"
plot(cars)

#ggplot resized to 2" x 2"
library(ggplot2)
ggplot(cars) +
  geom_point(aes(x = speed, y = dist))
```


### Using code chunks for non-code-like output

- must explicitly load `library("knitr")`
- Use `kable()` to produce pretty tables in pdf and Word output.
- `kable()` tables are less pretty in html output (including in-line output and Preview output)
- show example notebook

- Use `include_graphics()` to add already existing pictures.  **NOTE:** figure sizes don't affect these.  You have to use the `out.width` chunk option which can accept quoted values like `"2in"` or `"50%"`.  Unfortunately, `out.width` doesn't affect in-line output or Preview output, only knit output.



## YAML header
- Defaults
- Many other options including, but not limited to:
     + subitle
     + author
     + date
     + many output-specific options

## Advanced options for reports

- In gear icon, output options...
- Code-folding
- themes and highlighting
- Floating TOC with html. Tutorial [here](http://rmarkdown.rstudio.com/html_document_format.html#table_of_contents)

   
